---
title: "R features relevant for debugging"
output: html_document
---

# R features relevant for debugging {#features}

## Interpreter

R is an interpreted language, which means that code is interpreted by the R
interpreter during runtime. This has the effect that syntax errors occur only
when the code is run. Thus using an IDE like Rstudio makes coding easier as it
notifies the user on such errors beforehand.

When R exists it also can save the current environment (aka. workspace) into a
`.RData`-file. One should be careful to make certain that existing workspace does
not interfere with the code.

For non-interactive usage one usually uses `Rscript` or `R CMD BATCH`. `Rscript`
is usually better. However, when interpreter is launched using `Rscript`, the list
of default packages is minimal:
```{sh}
Rscript --help
```


## Functions

### R is a functional language

R is at its core a functional language. This means that coding is often
done in terms of functions that operate on objects and which return a new
object. These function calls are then often chained together
and vectorized with ``apply``-family of functions.

For more info, see
[this chapter of Advanced R](https://adv-r.hadley.nz/fp.html).[@wickham2019advanced]

### Lazy evaluation

R uses lazy evaluation when functions are called.
This means that functions are evaluated only when the output is actually used.

In the following example the value of ``g(x)`` is not evaluated for positive values and
thus the undefined variable is not encountered until ``f(x)`` is called with a negative
value:

```{r,  collapse=TRUE, error=TRUE}
g <- function(x) {
  return(x + z)
}

f <- function(x,y = g(x)) {
  if (x>0) {
    return(x)
  } else {
    return(y)
  }
}
f(1)
f(-1)
```
Lazy evaluation can make it easy to miss bugs in alternative execution paths.

See this [r-bloggers' post](https://www.r-bloggers.com/2018/07/about-lazy-evaluation/)
for more information on lazy evaluation.

### R functions are evaluated in an environment

Each R function has an environment that houses known names to values
(e.g. function names, variables etc.).

The global environment is the top level environment that contains everything
done by the user. Function calls etc. get their own environment that inherit
their parent's environment.

When running R code interactively the global environment can often become 
filled with various variables and definitions. Thus it is a good idea to clean
up the environment every now and then and verify that your code works from a clean
slate.

Thus one can easily run into situations such as this:

```{r, collapse=TRUE, error=TRUE}
  f(-2)
  z <- 2
  f(-2)
```
Objects can also write to their parents' environment using the `<<-`-operator:

```{r, collapse=TRUE}
h <- function(x) {
  x <<- x
}
x <- 1
x
h(2)
x
```

This can obviously cause problems and should be avoided in most cases.

For more info, see
[this chapter of Advanced R](https://adv-r.hadley.nz/environments.html).

## Objects

### All objects in R have a base type

There are 25 base types in R such as ``integer``, ``double``, ``list`` and
``closure`` (function). One can check the type using ``typeof(x)``-function.

```{r, collapse=TRUE}
a <- 1:10
b <- list(a=1:10)
c <- data.frame(b)
typeof(a)
typeof(b)
typeof(c)
typeof(f)
```
R is strongly, but dynamically typed. This means that objects have to keep their type, but objects can be cast to new types automatically. As an example, consider the following function that does a simple division:
```{r, collapse=TRUE}
d <- function(x,y) {
  print(is.integer(x))
  print(is.integer(y))
  print(is.integer(x/y))
  return(x/y)
}
d(10,3)
d(10L,3L)
```

For more info, see
[this chapter of Advanced R](https://adv-r.hadley.nz/base-types.html).

### Vectors are everywhere and they are immutable

R is a heavily vectorized language. Most mathematical operations in R are sped up by
doing them to the vector as a whole.

You can think of vectors as a collection of a thing with a length.
There are two types of vectors in R: **atomic vectors** and **list vectors**.

Atomic vectors are your typical logical, numeric (integer or double) and character vectors.

```{r, collapse=TRUE}
x_vector <- 1:6
is.atomic(x_vector)
```

Each vector has a type and a length:
```{r, collapse=TRUE}
typeof(x_vector)
length(x_vector)
```

Lists are vectors that can contain other vectors. They are not atomic. Elements in a list can have names.
```{r, collapse=TRUE}
x_list <- list(x_vector=x_vector)
x_list
names(x_list)
is.vector(x_list)
is.atomic(x_list)
```

Whenever you modify objects R will copy the data to a new location:
```{r, collapse=TRUE}
x_vector2 <- x_vector
tracemem(x_vector) == tracemem(x_vector2)
x_vector2 <- x_vector2 + 1
tracemem(x_vector) == tracemem(x_vector2)
```

However, when you modify lists, only the list changes, but not the atomics it points to. This makes it excellent for storing data:
```{r,  collapse=TRUE}
x_list2 <- x_list
tracemem(x_list) == tracemem(x_list2)
x_list2$x_vector2 <- x_vector2
tracemem(x_list) == tracemem(x_list2)
tracemem(x_list$x_vector) == tracemem(x_list2$x_vector)
```

Most R objects are built on top of vectors or lists by giving them attributes.
When they get attributes, they cease to be pure vectors.

```{r, collapse=TRUE}
print(x_vector)
print(attributes(x_vector))
print(is.vector(x_vector))
x_array <- array(x_vector, dim=c(3,2))
print(x_array)
print(attributes(x_array))
print(is.vector(x_array))
```

`data.frame` is an extremely important object in R built on top of a list.
It has an additional constraints added to it:

- All vectors in a data frame need to have the same length.
- It has `rownames()` and `colnames()`. `names()` of the data frame are
  the column names.
- A data frame has `nrow()` rows and `ncol()` columns. The `length()`
  of a data frame gives the number of columns.

`tibble` is a modern drop-in replacement for `data.frame` that is highly
recommended for its usability.
For more information on tibbles, see
[the packages web page](https://tibble.tidyverse.org/).[@R-tibble]

`data.table` is another popular alternative for `data.frame`. It has its own
usability enhancements and it scales better for big data than a normal data frame. 
For more information on `data.table`, see
[the packages web page](https://tibble.tidyverse.org/).[@R-data.table]

For more info on vectors, lists and data frames, see
[this chapter of Advanced R](https://adv-r.hadley.nz/vectors-chap.html).

### Base types are extended using object-oriented (OO) programming

Examples in this section utilize the [sloop](https://cran.r-project.org/web/packages/sloop/index.html)-package
for finding more information on objects. You can install it with:
```{r, eval=FALSE}
install.packages("sloop")
```

In R nomenclature, everything is an object. But not everything is an OO-style object.
OO-objects have a `class`-attribute. One can also use `is.object()`-function to check
whether the object is an OO object and `sloop::s3_class()` to get the class in a reliable way:
```{r, collapse=TRUE}
a <- 1:10
b <- list(a=1:10)
c <- data.frame(b)
attr(a, "class")
attr(b, "class")
attr(c, "class")
is.object(a)
is.object(b)
is.object(c)
sloop::s3_class(a)
sloop::s3_class(b)
sloop::s3_class(c)
```
One can strip the object back to its base types with `unclass()`-function:
```{r, collapse=TRUE}
unclass(c)
```

There are multiple OO paradigms in R:
  - [S3](https://adv-r.hadley.nz/s3.html): Oldest and simplest system. Most of R OO-objects are S3 objects.
  - [S4](https://adv-r.hadley.nz/s4.html): More advanced version of S3. Heavily used by the [Bioconductor](https://www.bioconductor.org/)-project.
  - [R6](https://adv-r.hadley.nz/r6.html): More like other OO languages. Improved version on R's feature called [reference classes](https://www.rdocumentation.org/packages/methods/versions/3.6.2/topics/ReferenceClasses). See [R6 manual for more information](https://r6.r-lib.org/)[@R6manual].

### Method dispatch

Most R functions are S3 **generic** functions that choose between available **methods** for a class. One can check whether a function is a generic function or a method with `sloop::ftype()`.

For example, `paste()` is an internal method that works with types, but `print()` is an S3 generic:
```{r, collapse=TRUE}
sloop::ftype(paste)
sloop::ftype(print)
```
To see which function is used for a given object, one can use `sloop::s3_dispatch()`:
```{r, collapse=TRUE}
sloop::s3_dispatch(print(c))
```
Now one can check the difference between these methods:
```{r, collapse=TRUE}
sloop::ftype(print.data.frame)
sloop::ftype(print.default)
print.data.frame(c)
print.default(c)
```
Knowing about method dispatch is especially important when dealing with numerical data. Numeric can mean that something is a double **or** that something behaves like a number (integer and double). Typically everything in R is stored as doubles.
```{r, collapse=TRUE}
x_i <- 1L
x_n <- 1
is.integer(x_i)
is.numeric(x_i)
is.integer(x_n)
is.numeric(x_n)
```
Another problem that can happen is that one can easily remove the class functionality from an S3 object. Let's consider a simple linear model. After fitting a linear model, one might try to store the coefficients into a variable:
```{r, collapse=TRUE}

x <- 1:10
err <- rnorm(10, 0, 0.5)
y <- 10 * x - 10 + err
d <- data.frame(x=x, y=y)

my_model <- lm(y ~ x, data=d)

my_coefs <- my_model$coefficients

my_coefs
```
After this, one might do a predictive fit:
```{r, collapse=TRUE}
y_pred <- my_coefs[2] * x + my_coefs[1]
y_pred
```
However, coefficients do not contain all of the information that the model has. Instead of ripping the coefficients out from the object, one should utilize the S3 generic function `predict` for models that support it. This will allow the model to stay as an S3 object:
```{r, collapse=TRUE}
attributes(my_model)

sloop::ftype(predict)
predict(my_model)
```
Utilizing these generic functions gives R great power, as one can use similar coding structures for various different models. For example, `predict` can operate on multiple different models with corresponding methods: 
```{r}
methods(predict)
```

## Signals and error handling

R has a robust system of messages, warnings and errors, that allow users
to prevent erroneous behaviour.

R has three different signal types:

- Messages: Messages are meant for informing the user that some action has been taken.
- Warnings: Warnings are meant to signal that not everything went correctly, but
  the program execution will continue.
- Errors: Errors indicate to R that something went wrong and the program execution should
  stop.

To raise these signals, one can use the following functions:
```{r, collapse=TRUE, error=TRUE}
message('This is a message')
warning('This is a warning')
stop('This is an error')
```

One can suppress these messages with `try()`, `suppressWarnings()` and `suppressMessages()`:
```{r, error=TRUE, collapse=TRUE}
signalsender <- function(signal_function, text) { signal_function(text) }
signalsender(message, 'This is a message')
suppressMessages(signalsender(message, 'This is a message'))

signalsender(warning, 'This is a warning')
suppressWarnings(signalsender(warning, 'This is a warning'))

try(signalsender(stop, 'This is an error'))
try(signalsender(stop, 'This is an error'), silent=TRUE)
signalsender(stop, 'This is an error')
```
By default `try` will still print that an error has occured, but it can be silenced.
However, silencing errors is risky and should only be used when the reason for the error is
known and expected.

All of the signals have handlers: functions that react when the signals are raised.
These handlers can be overwritten with `tryCatch()`- and `withCallingHandlers()`-functions:
```{r, collapse=TRUE, error=TRUE}
tryCatch(
  error = function(cond) {
    message(paste('Received an error\n', cond))
  },
  signalsender(stop, 'Raise an error')
)
withCallingHandlers(
  message = function(cond) {
    stop(paste('I did not expect a message:\n', cond,'\nRaising an error!'), call.=FALSE)
  },
  signalsender(message, 'Send a message')
)
```
There's a difference between these two functions. From Advanced R:

> - `tryCatch()` defines exiting handlers; after the condition is handled, control returns to the context where tryCatch() was called. This makes tryCatch() most suitable for working with errors and interrupts, as these have to exit anyway.
  - `withCallingHandlers()` defines calling handlers; after the condition is captured control returns to the context where the condition was signalled. This makes it most suitable for working with non-error conditions.

The difference is also well illustrated in the following quote from Advanced R:

> - An exiting handler handles a signal like you handle a problem; it makes the problem go away.
  - A calling handler handles a signal like you handle a car; the car still exists.

[rlang](https://rlang.r-lib.org/)-package[@R-rlang] provides additional wrappers for these signals that
can make error raising and handling easier. You can easily create your own custom error types with it.

For more info on R's error handling and on using `rlang`, see
[this chapter of Advanced R](https://adv-r.hadley.nz/conditions.html).

## Libraries

### Libraries are usually compiled

R libraries can be pure R or they can utilize other languages. Many R's internal functions
utilize R's [API for writing extensions](https://cran.r-project.org/doc/manuals/R-exts.html).
Common used languages are C, C++ and Fortran. Reason for this is speed provided by these lower´
level languages.

However, this API can be very complicated and thus most new packages use C/C++ with
[Rcpp](https://cran.r-project.org/web/packages/Rcpp/index.html)-package.[@R-Rcpp]

This means that during installation many R libraries need C and C++ compilers and external libraries,
which can result in various errors if you lack said requirements.

Debugging C or C++ code requires one to utilize C/C++ debuggers. For more information one
can look for example at
[this blog post](https://blog.davisvaughan.com/2019/04/05/debug-r-package-with-cpp/)[@VaughanRcppDebug] or
at the [various links](https://adv-r.hadley.nz/debugging.html#debug-compiled) provided in Advanced R.

For more information on Rcpp, one can check [Rcpp for everyone](https://teuder.github.io/rcpp4everyone_en/)[@RcppForEveryone], [this chapter in Advanced R](https://adv-r.hadley.nz/rcpp.html)