---
title: "Execises"
output: html_document
---

# Exercises

## Exercise on interactive debugging

Let's try out interactive debugging with a simple machine learning problem: using k-means clustering on the famous iris-dataset.

In this exercise there are four bugs hidden in the code.
Try to use debugging tools presented in [the debugging chapter](#debugging) to spot them out.
The bugs should appear in order, so if you fix one, you can continue to the next. 

This example is available here: [exercise-k-means.R](https://raw.githubusercontent.com/AaltoSciComp/r-debugging/main/examples/exercise-k-means.R).
Hints for each of the problems are available here:
[exercise-k-means-hints.txt](https://raw.githubusercontent.com/AaltoSciComp/r-debugging/main/examples/exercise-k-means-hints.txt).
Solution is available here:
[exercise-k-means-solution.R](https://raw.githubusercontent.com/AaltoSciComp/r-debugging/main/examples/exercise-k-means-solution.R.R).

```{r, collapse=TRUE, error=TRUE}
normalize_standard <- function(dataset) {
  # Normalizes each column in the dataset to standard normalization
  # https://en.wikipedia.org/wiki/Standard_score
  normalized_data <- data.frame()
  for (column in colnames(dataset)) {
    data_mean <- mean(dataset[[column]])
    data_sd <- sd(dataset[column])
    normalized_data[[column]] <- ((dataset[[column]] - data_mean))/data_sd
  }
  return(normalized_data)
}

kmeans_iris <- function() {
  # Load data
  data(iris)
  
  # Get a copy of the iris dataset
  iris_data <- iris
  
  # Normalize data
  iris_normalized <- normalize_standard(iris_data)
  
  # Take species out
  species <- iris_data$Species
  iris_data$Species <- NULL
  
  # Calculate kmeans clustering
  # https://en.wikipedia.org/wiki/K-means_clustering
  clustering <- kmeans(iris_normalized, 3)
  
  # Calculate confusion matrix
  # https://en.wikipedia.org/wiki/Confusion_matrix#Confusion_matrices_with_more_than_two_categories
  table(species, clustering$clusters)
}

print(kmeans_iris())
```

## Exercise on non-interactive debugging

Let's try non-interactive debugging with another simple problem: calculating surviability among Titanic passengers based on different features.

In this exercise there are two bugs hidden in the code. One of the bugs is a traditional
error raising bug, but the other one is a functional one.
Try to use debugging tools of the [non-interactive debugging](#noninteractive) section to
find out the bugs.

This example is available here: [exercise-titanic-survivability.R](https://raw.githubusercontent.com/AaltoSciComp/r-debugging/main/examples/exercise-titanic-survivability.R).
Hints for each of the problems are available here:
[exercise-k-means-hints.txt](https://raw.githubusercontent.com/AaltoSciComp/r-debugging/main/examples/exercise-titanic-survivability-hints.txt).
Solution is available here:
[exercise-titanic-survivability-solution.R](https://raw.githubusercontent.com/AaltoSciComp/r-debugging/main/examples/exercise-titanic-survivability-solution.R).



```{r, collapse=TRUE, error=TRUE}
# Load titanic survivability data
data(Titanic)

titanic <- as.data.frame(Titanic)

check_survivability <- function(feature) {
  # Function for checking survivability per feature
  
  # Get the different feature levels
  levels <- levels(as.factor(titanic[[feature]]))
  
  # Create output data frame
  titanic_survivability <- data.frame(row.names=levels)
  
  # Calculate survivability per level
  for (level in levels) {
    level_data <- titanic[Titanic[[feature]] == level, ]
    n_survived <- sum(level_data[level_data$Survived == 'Yes', 'Freq'])
    n_died <- sum(level_data[level_data$Survived == 'No', 'Freq'])
    titanic_survivability[levels, 'survivability'] <- n_survived / (n_survived + n_died)
  }
  return(titanic_survivability)
}

# Features to iterate over
different_features <- c('Class', 'Sex', 'Age')

# Analyze survivability per feature
for (feature in different_features) {
  cat(paste('Feature to check:', feature, '\n'))
  print(check_survivability(feature))
}
```

