--- 
title: "R debugging"
author: "Simo Tuomisto"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
csl: chicago-fullnote-bibliography.csl
#link-citations: yes
description: "Materials for Aalto Scientific Computing's How to debug-course, R part"
output: html_notebook
---

# About this course

This course is a part of
[How to debug code](https://scicomp.aalto.fi/training/scip/debugging-2022/)-course
by Aalto Scientific Computing.

The course structure and some of the material is adapted from the
[Python part of the debugging course](https://aaltoscicomp.github.io/python-debugging).

Most of the R material is constructed from various books and blog posts available online.
Reading this material for more in-depth explanation of concepts shown in this course
is highly recommended. [Advanced R by Hadley Wickham](https://adv-r.hadley.nz/)[@wickham2019advanced]
especially cannot be recommended enough.

## Requirements

This course utilizes [RStudio IDE](https://www.rstudio.com/) as it provides easy access
for many debugging tools.

[sloop](https://cran.r-project.org/web/packages/sloop/index.html)-package[@R-sloop] is used
to check some R object features:
```{r, eval=FALSE}
install.packages("sloop")
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'sloop', 'Rcpp'
), 'packages.bib')
```
